Joint maximum-likelihood of phylogenies and ancestral states is not consistent

### 2018-06-22 appendix.tex
condition on gradients and plot of intuition

@@

\begin{figure}
\centering
\includegraphics[width=.95\textwidth]{empirical-incons}
\caption{
Inconsistency in joint inference.
In the white region, one of the possibly many equally likely maxima has the property that $\hat{w}=1$ and $\hat{y}_2=1$, indicating we converge to a topology that is not the one that generated the data.
The gray regions are edge cases where one of the generating parameters was zero or one.
}
\label{fig:empirical-incons}
\end{figure}

>>>>

Tentative lemma:
\begin{lemma}
Let $t^*=\{x^*, y^*, x^*, y^*, y^*\}$ and $t=\{x_1, y_1, x_2, y_2, w\}$ and $\emptyset$ be the maximal ancestral state for all site splits.
There exists a set of $0 < x^*, y^* < 1$ such that $\{x^*(y^*)^2, (y^*)^2, x^*(y^*)^2, 1, 1\}$ is the solution to maximizing \eqref{eq:profile_likelihood}.
\end{lemma}

Sketch of proof:
We know the gradient is zero for the partials of the first three parameters.
We must show then that the objective function is increasing in all directions for the second two parameters at the boundary.
If the gradients for both $w$ and $y_2$ are positive at 1, the objective will be increasing in all increasing directions---for any positive direction, i.e., unit vector with nonnegative entries, the directional derivative is positive if each gradient is positive.
Since the objective function is concave then this must be the maximum.
Let
\[
a_{\emptyset} = \frac{(x_1+x_2)(y_1+y_2)}{(1+x_1x_2)(1+y_1y_2)}
\]
with similar definitions for the remaining site splits.
The objective function proportional to $w$ is
\begin{align*}
\ell(w) \propto \log(1+w) + \sum_j p_{\siteSplit_j} \log(1 + a_{\siteSplit_j}w)
\end{align*}
and has gradient with respect to $w$ as
\begin{align*}
\ell_w(t) = \frac{1}{1+w} + \sum_j p_{\siteSplit_j}a_{\siteSplit_j}\frac{1}{1 + a_{\siteSplit_j}w}.
\end{align*}
Note that, $a_{\emptyset} = -a_{13}$ and $a_{2} = -a_{123}$, and, for $x_1=x_2$,
\[
a_{1} = a_{3} = a_{12} = a_{23} = 0.
\]

For $y_2$, we do the same, but now with
\begin{align*}
\ell(y_2) \propto \log(1+y_2) + \sum_j p_{\siteSplit_j} \log(1 + b_{\siteSplit_j}y_2)
\end{align*}
where
\[
b_{\emptyset} = \frac{y_1+wx_1+wx_2+x_1y_1x_2}{1+x_1x_2+wx_1y_1+wx_2y_1}.
\]
For $x_1=x_2$, the denominator is always positive, allowing the above simplification.
%It is. Since x_1=x_2 at the maximum we can use an argument like 1+x^2-2y_1wx > 1+x^2-2x = (1-x)^2 > 0. We'll need to ignore cases where x=0 (i.e., x^* or y^* are zero)
The constants $b$ do not simplify as easily.
In Fig.~\ref{fig:positive-gradients}, we see the region where our solution is $\hat{t}=\{x^*(y^*)^2, (y^*)^2, x^*(y^*)^2, 1, 1\}$.

\begin{figure}
\centering
%python joint_inf_plot.py --analytic --plot-name svg-fig/positive-gradients.svg --delta .001
\includegraphics[width=.95\textwidth]{positive-gradients}
\caption{
}
\label{fig:positive-gradients}
\end{figure}

@@

\begin{figure}
\centering
%python joint_inf_plot.py --empirical --plot-name svg-fig/empirical-incons.svg --delta .01 --n-jobs 10
\includegraphics[width=.95\textwidth]{empirical-incons}
\caption{
Inconsistency in joint inference.
In the white region, one of the possibly many equally likely maxima has the property that $\hat{w}=1$ and $\hat{y}_2=1$, indicating we converge to a topology that is not the one that generated the data.
The gray regions are edge cases where one of the generating parameters was zero or one.
}
\label{fig:empirical-incons}
\end{figure}

\begin{figure}
\centering
%python joint_inf_plot.py --empirical --plot-name svg-fig/empirical-incons-curve.svg --delta .01 --n-jobs 10 --plot-curve
\includegraphics[width=.95\textwidth]{empirical-incons-curve}
\caption{
Inconsistency in joint inference with intuitive curves.
The white curve is our condition on the gradients.
The blue curve is a condition on $x^*$ and $y^*$ such that $\emptyset$ is the most likely ancestral state.
}
\label{fig:empirical-incons}
\end{figure}
